---
title: "Irrigation Analysis GSSM25"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_notebook: default
  html_document:
    toc: true
    number_sections: true
    df_print: paged
  pdf_document: default
params:
  test_mode_active: FALSE
  required_packages: ["readxl", "writexl", "dplyr", "purrr", "tibble", 
                      "tidyr", "ggplot2", "stringr", "maps"]
  root_path: "D:/OneDrive - USDA/Desktop/GSSM25 Polished for Distribution"
  Data_Source: "USDA NASS Census"
  census_tables: ['irrigation', 'overview']
  selected_states: ['SC', 'NC', 'GA']
  file_path_irrigation_SC: "/st45_010.xlsx"
  file_path_irrigation_NC: "/st37_010.xlsx"
  file_path_irrigation_GA: "/st13_010.xlsx"
  file_path_overview_SC: "/st45_1.xlsx"
  file_path_overview_NC: "/st37_1.xlsx"
  file_path_overview_GA: "/st13_1.xlsx"
  columns_irrigation: ['Irrigated farms (number)', 
                       'Land in irrigated farms (acres)']
  columns_overview: ['Number of Farms', 'Acreage in Farms']
  percentages: ['percentage_number', 'percentage_acreage'] 
  percentage_number: ['Irrigated farms (number)', 'Number of Farms']
  percentage_acreage: ['Land in irrigated farms (acres)', 'Acreage in Farms']
  percentages_names: 
     ['Farms with Irrigation (%)', 'Irrigated Farm Acreage (%)'] 
---

# Introduction

This R markdown file uses data in the associated excel files to generate 
plots of interest.

New data can be added by simply processing the data, then altering the YAML
header at the top of the file.

# Libraries

This code loads the following libraries, which are listed in the YAML header.

```{r, import_libraries, echo = FALSE, warning=FALSE, message=FALSE}

  invisible(lapply(params$required_packages, library, character.only = TRUE))
  print(params$required_packages) # list the libraries installed

```

# Data Processing Steps

The data processing (or data wrangling) steps are the initial steps for 
importing data from the excel files (or simulating data) for output

## Import Data from Excel

```{r, import_data, echo=FALSE}

if (!params$test_mode_active){  # Note: This code will not run in test mode

  # Load data from excel spreadsheets
  NASS_data <- 
    lapply(params$census_tables, function(target_table) {
     lapply(params$selected_states, function(target_state) {
        
        message("Importing ", target_table, " for ", target_state) 
        
        # Get file path and column information from YAML parameters
        file_param <- paste0('file_path_', target_table, '_', target_state)
        target_file_path <- paste0(params$root_path, params[[file_param]])
        target_columns <- params[[paste0("columns_", target_table)]] 

        lapply(excel_sheets(path = target_file_path), function(target_sheet)
        { 
          
          # extract the target year from the sheet name
          target_year <- sub(".*([12]\\d{3}).*", "\\1", target_sheet) %>% 
            as.integer()
          
          read_excel(path = target_file_path, # import the Excel file
                     sheet = target_sheet,
                     col_types = "text") %>%
            
            # tidy the imported column names
            mutate(Item = Item  %>%
                     gsub(",", "", .) %>%  # Remove commas
                     
                     # Add space before "acres"
                     gsub("(\\S)(acres)", "\\1 acres", .) %>%  
                     
                     # add parenthesis around acre or number
                     gsub("\\b(acres|number)\\b", "(\\1)", .) %>%
                     
                     gsub("\\s*\\d{4}.*$", "", .) %>%  # Remove the year
                     
                     gsub("\\(see text\\)", "", .) %>%  # Remove "(see text)"
                     
                     trimws() # trim white space
            ) %>% 
            
            # Remove the state level data from county level table
            select(-any_of(state.name)) %>%  
            
            # filter to keep target columns
            filter(Item %in% target_columns) %>%
            
            # transpose (pivot) the table so rows and columns swap from the 
            # Excel sheet to the R dataframe. 
            pivot_longer(-1) %>%
            pivot_wider(names_from = 1, values_from = value) %>%
            
            rename(county = name) %>% # label the county column
            
            # add state and year information in front of county
            mutate(state = target_state, year = target_year) %>%  
            relocate(state, year, .before = county) %>%  
            
            # Convert text imported from Excel to suitable R column data
            mutate(across(
              all_of(target_columns), 
              function(raw_text_column) { 
                sapply(raw_text_column, function(raw_text){
                  raw_text <- raw_text %>% 
                    trimws() %>% # trim white space
                    str_replace_all(" ", "") %>% # remove all spaces
                    
                    # '-' is equivalent to 0 in NASS data
                    str_replace_all('-', '0')
 
                  # if the data is numeric then return an integer
                  if (grepl("^[0-9]+$", raw_text)) 
                    return(as.integer(raw_text))
                  
                  # Handle the (D) case as NA which means too little data 
                  if (raw_text == '(D)' || raw_text == '') 
                    return(NA_integer_)                  
                 
                  return(NA_integer_)  # default to NA   
                }, USE.NAMES = FALSE)
              }))
        })
     }) %>% bind_rows()
    }) %>% reduce(full_join, by = c("state", "year", "county")) %>%
    
    # Calculate percentages  
    cbind(sapply(params$percentages_names, function(p) {
      p_list <- params$percentages[match(p, params$percentages_names)]
      numerator <- .[[params[[p_list]][1]]]
      denominator <- .[[params[[p_list]][2]]]
      
      return(
        ifelse(denominator != 0, 
               round(100 * numerator / denominator, 2), NA)
      )
    }, USE.NAMES = TRUE)) 
  
  # add column names as attribute for easy referencing later
  attr(NASS_data, "column_names") <- 
    params[str_detect(names(params),"^(columns_)")] %>% 
    unlist() %>% unique() %>% 
    c(params$percentages_names)
  
} 

```

## Test mode (if the data is not available) 

In the event no data is available an alternative testing mode can be set to
simulate the data so graphs can be worked on independent of the availability
of data.

```{r setup_test_mode, echo=FALSE}

if (params$test_mode_active) {
  
  Test_Environment <- new.env()
  with(Test_Environment, {
    DATE_RANGE <- 1978:2022
    NUM_YEARS <- 7
    VALUE_RANGE <- 0:10000  
    simulated_census_table <- params$census_tables
    simulated_states <- params$selected_states
    sim_state_names <- tolower(
      state.name[match(simulated_states, state.abb)])
    test_counties <- 
      subset(map_data("county"), region %in% sim_state_names) %>% 
      select(region, subregion) %>% unique()
    Simulate_States <- rep(test_counties$region, NUM_YEARS)
    Simulate_Counties <- str_to_title(
      rep(test_counties$subregion, NUM_YEARS))
    Simulate_Years <- rep(sample(DATE_RANGE, NUM_YEARS), 
                          each = length(test_counties$subregion))
    Random_Data <- function() {
      sample(VALUE_RANGE, nrow(test_counties) * NUM_YEARS, replace = TRUE)
    }    
  })
  
  NASS_data <- 
    lapply(Test_Environment$simulated_census_table, function(census_table) {
      lapply(Test_Environment$sim_state_names, function(selected_state) {
        
        simulated_columns <- params[[paste0("columns_", census_table)]]
        simulated_data <- data.frame(
          state = Test_Environment$Simulate_States,
          county = Test_Environment$Simulate_Counties,
          year = Test_Environment$Simulate_Years
        )
        for (col_name in simulated_columns) {
          simulated_data[[col_name]] <- Test_Environment$Random_Data()
        }
        
        return(simulated_data)
      }) %>% bind_rows()
    }) %>% reduce(full_join, by = c("state", "year", "county")) %>%
    
   cbind(sapply(params$percentages, function(percentage) {
       round(100 * 
               .[[params[[percentage]][1]]] /
               .[[params[[percentage]][2]]], 2) 
   }, USE.NAMES = TRUE))
  
  # Add column_names as an attribute to the simulated NASS_data
  
  attr(NASS_data, "column_names") <- 
    params[str_detect(names(params),"^(columns_)")] %>% 
    unlist() %>% unique() %>% 
    c(params$percentages_names)
  
}else{
  print("Test mode is inactive.")
}
```
\newpage

# Generate tables and plots

<!-- ## Top 25 Counties List -->

<!-- This is the list of top 25 counties for every variable. -->

```{r Top_25_Counties, results='asis', echo=FALSE}
# Top 25 counties sorted

YEAR <- as.integer(2022)

# function to sort the top data points
top_sorted_data <- function(column_name, year_int) {
  NASS_data %>%
    filter(year == year_int) %>%
    mutate(county_state = paste(county, state, sep = ", ")) %>%
    select(county_state, year, all_of(column_name)) %>%
    arrange(desc(as.numeric(.data[[column_name]]))) %>%
    head(25)
}

# lapply(attr(NASS_data, "column_names"), function(column_name){
# 
#   print(knitr::kable(
#     top_sorted_data(column_name, YEAR) %>% select (-year), "pandoc",
#     caption = paste("Top 25", column_name, "in", YEAR), longtable = FALSE
#   ))
# }) %>% invisible ()


```

## Bar Graph of Top 25 Counties per Statistic

```{r Top_25_Bar_Graph, echo=FALSE}

YEAR <- as.integer(2022)

sapply(attr(NASS_data, "column_names"), function(column_name){
  
  # Create the bar graph for the top 25 counties
  print(
    ggplot(data = top_sorted_data(column_name, YEAR), 
           aes(x = reorder(county_state, !!sym(column_name)),  
               y = !!sym(column_name))) +
      geom_bar(stat = "identity", fill = "steelblue") +
      geom_text(aes(label = 
                      top_sorted_data(column_name, YEAR)[[column_name]]), 
                hjust = -0.3, 
                size = 3.5) +
      labs(
        title = paste(
          "Top 25 Counties in", 
          paste(params$selected_states, collapse = ", "),
          "in", YEAR
        ), 
        x = "County, State",
        y = column_name) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, size = 12)  # Center the title
      ) +
      coord_flip() +
      
      # Avoid cutoff of y-axis
      ylim(0, max(top_sorted_data(column_name, YEAR)[[column_name]]) * 1.2) 
  )
}) %>% invisible()

```

\newpage

## Box and Whiskers Plot

This is a box and whiskers plot to look at the distributions of county level
variables by state.

```{r Box_and_Whiskers_Plot, echo=FALSE, warning=FALSE}

# The year is fixed
YEAR <-  as.integer(2022)

sapply(attr(NASS_data, "column_names"), function(column_name){
  
  # Create a box and whisker plot for the current column and year
  print(ggplot(
    data = filter(NASS_data, year == YEAR), 
    aes(x = state, y = !!sym(column_name), fill = state)
  ) +
    geom_boxplot() +
    ggtitle(paste("Distribution of", 
                  column_name, 
                  "per County in", YEAR)) +
    labs(x = "State", y = column_name) +
    theme_minimal() +
    theme(
      legend.position = "none", # Hide the legend
      plot.title = element_text(size = 9, hjust = 0.5)  # Center the title
    ))
  
}) %>% invisible()
```

\newpage

## Region of Interest Map

This is a map of the region of interest.

```{r region_of_interest, echo=FALSE}
# This is to verify the map is printing

# make map to verify region of interest
make_mappable <- function(NASS_data){
  NASS_data %>% 
    mutate(county = tolower(county), 
           state = tolower(state.name[match(state, state.abb)]) )
}


# Plot the map with county borders 
print(
  ggplot(data = map_data("state") %>% # Get state boundary points
           filter(region %in% unique(make_mappable(NASS_data)$state)), 
         mapping = aes(x = long, y = lat, group = group)) + 
    coord_fixed(1.3) + 
    geom_polygon(color = "black", fill = "gray") + 
    geom_polygon(
      data = map_data("county") %>% # get count boundary points
        filter(region %in% unique(tolower(make_mappable(NASS_data)$state))),
                 fill = NA, color = "white") + 
    geom_polygon(color = "black", fill = NA) + 
  # ggtitle('Region of Interest') + 
  theme_void() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold")
  )
)

```

\newpage

## Choropleth Map

The choropleth map is to show the distribution of the county-level data 
according the the quartile the data.

```{r chloropleth_map, echo=FALSE, warning=FALSE}

YEAR <- 2022

# function to make a NASS data align with the mapping format
make_mappable <- function(NASS_data){
  NASS_data %>% 
    mutate(county = tolower(county), 
           state = tolower(state.name[match(state, state.abb)]) )
}

# get county boundary points and attach NASS data
county_boundary_points <- map_data("county") %>% 
  filter(region %in% unique(tolower(make_mappable(NASS_data)$state)))  %>%
  mutate(id = paste(region, subregion, YEAR, sep = "_")) %>% 
  left_join(make_mappable(NASS_data) %>% 
              mutate(id2 = paste(state, county, year, sep = "_")),
            by = c("id" = "id2"))

sapply(attr(NASS_data, "column_names"), function(column_name){
  
  # Calculate quartile and round the breaks for better readability
  quartiles <- quantile(county_boundary_points[[column_name]], 
                        probs = seq(0, 1, by = 0.25), na.rm = TRUE)
  
  
  # Create and plot the choropleth map with quartile-based fill, filtering by year
  print( 
    ggplot(data = county_boundary_points,  
           mapping = aes(x = long, y = lat, group = group, 
                         fill = cut(
                           .data[[column_name]],
                           breaks = quartiles,
                           labels = paste0("Q", 1:4, ": ", 
                                           round(quartiles[-5], 2), " - ", 
                                           round(quartiles[-1], 2)),
                           include.lowest = TRUE
                         )
           )) +
      coord_fixed(1.3) +  # Ensure aspect ratio is fixed
      geom_polygon(color = "black") +  # Draw polygons with black borders
      
      # Map the colors to the quartiles (Blue, Orange, Green, Red)
      scale_fill_manual(values = c("#1f77b4", "#2ca02c", "#ff7f0e",  "#d62728"), 
                        na.value = "grey50", 
                        drop = FALSE) + 
      
      # Add title with year
      ggtitle(paste0("Choropleth Map of ", 
                     column_name, 
                     " in ", 
                     YEAR)) +  
      labs(fill = "Quartile (Range)") +  # Label the fill legend
      theme_minimal() +  # Apply a minimal theme
      theme(axis.text = element_blank(),  # Remove axis text
            axis.ticks = element_blank(),  # Remove axis ticks
            axis.title = element_blank())  # Remove axis titles
  )

}) %>% invisible()
```

\newpage

## Time Series Box and Whisker Plot

The time series box and whisker plot is to Visualize the distribution of the 
variables across years the data was recorded.

```{r time_series_bx_and_wskr_plot, echo=FALSE, message=FALSE, warning=FALSE}

# Loop through each column and generate the time series boxplot
sapply(attr(NASS_data, "column_names"), function(column_name){
  
  print(
    ggplot(NASS_data)   + 
      geom_boxplot(aes(x = reorder(year, state), 
                       y = .data[[column_name]], 
                       fill = year)) +
      ggtitle(paste("Time Series Boxplot of", 
                    column_name, 
                    "by Year in",
                    paste(params$selected_states, collapse = ", "))) +
      labs(x = "Year", y = paste(column_name, "per County")) +
      theme_minimal() +
      theme(
        legend.position = "none",
        plot.title = element_text(size = 9, hjust = 0.5)
      )
  )
}) %>% invisible()
```
